### 2024-09-25 teaching/math240f24/draft.tex
lecture sept 24

\medskip\boldlabel\datestamp{8}{} Injective and surjective functions.
The {\it range} or {\it image} of a function $f:A\to B$ is
$$f(A) = \bigl\{b\in B : \hbox{there exists}\ a\in A\ \hbox{such that}\ b = f(a)\bigr\}.$$
These are all the values $f$ actually outputs. For instance, if we let $f : \ZZ\to \NN$
be given by $f(n) = n^2$, then
$$\eqalign{
f(\ZZ) &= \{\ldots, f(-2), f(-1), f(0) , f(1), f(2),\ldots\} \cr
&= \{\ldots, 9, 4, 1, 0 1, 4, 9,\ldots\} \cr
&= \{0, 1, 4, 9, 16, 25, \ldots\}. \cr
}$$
A function $f:A\to B$ is called {\it surjective} or {\it onto} if $f(A) = B$, that is,
for every $b\in B$ there exists some $a\in A$ such that $f(a) = b$. An example of a
surjective function is $f:\QQ\to \QQ$ sending $x\mapsto x/2$. This is because for any $q\in \QQ$,
we can set $r = 2q$, and
$$f(r) = {r\over 2} = {2q\over 2} = q.$$
A function $f:A\to B$ is called {\it injective} or {\it one-to-one} if for all $a_1,a_2\in A$
with $a_1 \ne a_2$, we also have $f(a_1)\ne f(a_2)$. Equivalently, $f$ is injective if
$f(a_1) = f(a_2)$ implies that $a_1 = a_2$ for all $a_1, a_2 \in A$. For instance, the function
$f:\ZZ\to \NN$ that sends $n\mapsto n^2$ is not injective since $f(-1) = f(1)$, but $-1 \ne 1$.
On the other hand, if we modify the domain, considering $f:\NN\to \NN$ sending $n\mapsto n^2$,
then now $f$ is injective, since if $f(m) = f(n)$, then $m^2 = n^2$, and there is only one positive
integer that squares to any given integer, so $m=n$.

@@

\proclaim Theorem \advthm. Let $a_1, a_2, \ldots, a_m$ be a finite sequence (repeats allowed) of
real numbers. Let
$$a = {1\over m}\sum_{i=1}^m a_i$$
be the average value of the sequence and let $m$ be the maximum value the sequence attains. Then $m \ge a$.

\proof We have
$$a = {1\over m}\sum_{i=1}^m a_i \le {1\over m} \sum_{i=1}^m m = {m^2\over m} = m.\noskipslug$$

@@

\proof For $1\le i \le n$, let $r_i$ be the number of $a\in A$
such that $f(a) = b_i$. This is a sequence that adds up to $m$, since every $a$ in $A$ maps to
exactly one element of $B$. So the sequence has average $m/n$, and by the previous theorem,
there must be some $j$ such that $r_j \ge m/n$. But the $r_i$ are all actually integers (since
cardinalities of finite sets are integers), meaning that $r_j \ge \lceil m/n \rceil$. Letting
$b = b_j$ completes the proof.\slug

@@

We define $g : F\to E$ by
$$g(A) = \cases{A\setminus\{x\}, & if $x\in A$;\cr A\cup \{x\}, & if $x\notin A$.\cr}$$
As before, $\bigl|g(A)\bigr|$ is even, since $A$ is assumed to be a member of $F$ now. Now
for any $A\in E$,
$$\eqalign{
g\bigl(f(A)\bigr) &= \cases{g\bigl(A\setminus\{x\}\bigr), & if $x\in A$;\cr
g\bigl(A\cup \{x\}\bigr), & if $x\notin A$\cr} \cr
&= \cases{\bigl(A\setminus\{x\}\bigr)\cup \{x\}, & if $x\in A$;\cr
\bigl(A\cup \{x\}\bigr)\setminus \{x\} , & if $x\notin A$\cr} \cr
&= \cases{A, & if $x\in A$;\cr
A, & if $x\notin A$\cr} \cr
&= A.\cr
}$$
The proof that $f\bigl(g(A)\bigr) = A$ is similar. Thus $g$ is the inverse of $f$.\slug

>>>>

\medskip\boldlabel\datestamp{24}{ix} Injective and surjective functions.
The {\it range} or {\it image} of a function $f:A\to B$ is
$$f(A) = \bigl\{b\in B : \hbox{there exists}\ a\in A\ \hbox{such that}\ b = f(a)\bigr\}.$$
These are all the values $f$ actually outputs. For instance, if we let $f : \ZZ\to \NN$
be given by $f(n) = n^2$, then
$$\eqalign{
f(\ZZ) &= \{\ldots, f(-2), f(-1), f(0) , f(1), f(2),\ldots\} \cr
&= \{\ldots, 9, 4, 1, 0 1, 4, 9,\ldots\} \cr
&= \{0, 1, 4, 9, 16, 25, \ldots\}. \cr
}$$
A function $f:A\to B$ is called {\it surjective} or {\it onto} if $f(A) = B$, that is,
for every $b\in B$ there exists some $a\in A$ such that $f(a) = b$. An example of a
surjective function is $f:\QQ\to \QQ$ sending $x\mapsto x/2$. This is because for any $q\in \QQ$,
we can set $r = 2q$, and
$$f(r) = {r\over 2} = {2q\over 2} = q.$$
A function $f:A\to B$ is called {\it injective} or {\it one-to-one} if for all $a_1,a_2\in A$
with $a_1 \ne a_2$, we also have $f(a_1)\ne f(a_2)$. Equivalently, $f$ is injective if
$f(a_1) = f(a_2)$ implies that $a_1 = a_2$ for all $a_1, a_2 \in A$. For instance, the function
$f:\ZZ\to \NN$ that sends $n\mapsto n^2$ is not injective since $f(-1) = f(1)$, but $-1 \ne 1$.
On the other hand, if we modify the domain, considering $f:\NN\to \NN$ sending $n\mapsto n^2$,
then now $f$ is injective, since if $f(m) = f(n)$, then $m^2 = n^2$, and there is only one positive
integer that squares to any given integer, so $m=n$.

@@

\proclaim Theorem \advthm. Let $a_1, a_2, \ldots, a_n$ be a finite sequence (repeats allowed) of
real numbers. Let
$$a = {1\over n}\sum_{i=1}^n a_i$$
be the average value of the sequence and let $m$ be the maximum value the sequence attains. Then $m \ge a$.

\proof We have
$$a = {1\over n}\sum_{i=1}^n a_i \le {1\over n} \sum_{i=1}^n m = {mn\over n} = m.\noskipslug$$

@@

\proof Enumerate $B = \{b_1, b_2, \ldots, b_n\}$. For $1\le i \le n$, let $r_i$ be the number of $a\in A$
such that $f(a) = b_i$. This is a sequence that adds up to $m$, since every $a$ in $A$ maps to
exactly one element of $B$. So the sequence has average $m/n$, and by the previous theorem,
there must be some $j$ such that $r_j \ge m/n$. But the $r_i$ are all actually integers (since
cardinalities of finite sets are integers), meaning that $r_j \ge \lceil m/n \rceil$. Letting
$b = b_j$ completes the proof.\slug

@@

We define $g : F\to E$ by
$$g(A) = \cases{A\setminus\{x\}, & if $x\in A$;\cr A\cup \{x\}, & if $x\notin A$.\cr}$$
As before, $\bigl|g(A)\bigr|$ is even, since $A$ is assumed to be a member of $F$ now. Now
for any $A\in E$,
$$\eqalign{
g\bigl(f(A)\bigr) &= \cases{g\bigl(A\setminus\{x\}\bigr), & if $x\in A$;\cr
g\bigl(A\cup \{x\}\bigr), & if $x\notin A$\cr} \cr
&= \cases{\bigl(A\setminus\{x\}\bigr)\cup \{x\}, & if $x\in A$;\cr
\bigl(A\cup \{x\}\bigr)\setminus \{x\} , & if $x\notin A$\cr} \cr
&= \cases{A, & if $x\in A$;\cr
A, & if $x\notin A$\cr} \cr
&= A.\cr
}$$
The proof that $f\bigl(g(A)\bigr) = A$ is similar. Thus $g$ is the inverse of $f$.\slug
\bye
